---
title: "Final Project"
author: "Levi Basom"
output: html_notebook
---

## Research Question

The reasearch question I am choosing to ask is:

**Could relevant firearm permit numbers in a year have some explanation for trends in violent crime data and hunting data?**

By choosing to ask this type of question, I will need to find AT LEAST 3 relevant data sets and think of ways to combine them to create graphs and statistics that help to answer parts of this question.

My first (and primary) data source will be the firarm permit data, because it will separately be related to my other two data sources. My second data source will be the hunting data, and my third data source will be the violent crime data in the United States.

I will look at how increases in certain types of firearm permits could relate to increases in hunting data, and how firearm permits for certain states can relate to violent crime within those states over certain periods of time, and if these fluctuations in both data tend to increase and decrease together, or mutually.

## Data Sources

### Load required packages

Before beginning my analysis, I need to load relevant packages that will allow my data to be loaded with the least problems and stored in a clean way, as well as ones that will give me the ability to create graphics and plots that show trends and patterns in the data. Packages that will help with this project (whether they will be used or not) are `ggplot2`, `tidyr`, `mosaic`, `DataComputing`, `readr`, and `data.table`.

```{r}
install.packages("data.table")
```

```{r}
library(ggplot2)
library(tidyr)
library(mosaic)
library(DataComputing)
library(readr)
library(data.table)
```

### Finding data sets

I will have to search for reputable and adequate data sources to match the three data sets that I intend to use in this project. After some time in searching with certain criteria in mind, I was able to find 3 sources. My primary data source was found from a routinely maintained Github repository that collects NICS firearm data over the years and constantly updates each month. My second source is Pennsylvania hunting data straight from the Pennsylvania Game Commission website, and my third data set is crime data for the 50 United States and the District of Columbia from the FBI Uniform Crime Reporting Statistics website.

#### Primary data:

The NICS Firearm Background Check Data is my primary data set for 3 main reasons:

1. The data encompasses all U.S. states and territories, making it easy to look at trends for individual states over the years.

2. The data has many variables relating to different types of guns, and includes many types of sales, like regular sales, private sales, returned sales, and many others.

3. The data breaks each year down into months and by state, so it is very specific and encompasses a large range of detail and information.

The data within this data set comes directly from the FBI National Instant Criminal Background Check System (NICS), and is collected and maintained from that website by Buzzfeed News with great documentation and reputability. 

The NICS data was collected to keep a very close estimate of firearm sales and possession per state and how that number has been increasing/decreasing over time. The data collection began (by the FBI) in **1998**, and has been consistenly collected and updated from 1998 until currently (**the latest data is from March of 2020**).

Each case in the data is uniquely defined by year, month, and state, so every case shows the count of certain types of permits for a certain state or territory in a given month of a year. There are **14,135** cases in the most updated state of the data set that I will be using for this project.

From all variables in this data set, I will be using the ones most relevant to either of the two secondary data sets that I will be using. There are three categories for guns in the NICS data: handgun, long gun, and other. For the hunting data, I will be using variables that relate to the most common types of guns used in hunting, like `long_gun`, `other`, `prepawn_long_gun`, `prepawn_other`, etc. For the crime data, I will most likely take a look at `handgun`, `prepawn_handgun`, etc. because handguns are the most common firearm used in crime, but I may also look into all types of firearms to see how the trends change.

Since this data set encompasses many less years than the other two data sets, I will most likely need to filter out a lot of the earlier rows in the other data sets in order to make relevant inferences when comparing data from two sets.

To load the data set from the Github URL, I will use `read_csv()` to create a `NICS` table:

```{r}
NICS_URL <- "https://raw.githubusercontent.com/BuzzFeedNews/nics-firearm-background-checks/master/data/nics-firearm-background-checks.csv"

NICS <- read_csv(NICS_URL)

head(NICS, 10)
```

In order to group by month and year more easily, we will need to clean the `NICS` data by splitting the `month` column into `Month` and `Year`, using `extractMatches()` and `rename()` to create these two new columns:

```{r}
NICS <- 
  NICS %>%
  extractMatches("(..$)", month) %>%
  rename("Month" = match1) %>%
  extractMatches("(^....)", month) %>%
  rename("Year" = match1)
```

```{r}
NICS %>%
  group_by(month, Month, Year) %>%
  select(month, Month, Year) %>%
  summarise(n()) %>%
  head(10)
```

As we can see from this summary, our data now has `Month` and `Year` columns that reflect a split in the `month` column (which follows the format "YYYY-MM").

#### Other data sources:

I was able to find a useful table on the Pennsylvania Game Commission website for hunting data in Pennsylvania. This data is collected and maintained by the PA Game Commission by measuring the number of tags for each killed animal that is reported (tag collected). The table on the website is titled "White-Tailed Deer, Bear, & Wild Turkey Harvest Information". The data contains information for years spanning **1915 through 2015**, so I am assuming that it has not been updated recently (but still relevant for the years that it contains). This data was collected with the goal of watching trends in hunting data for each provided species of animal over the years. A case in this data represents the amount of tags collected for each species of animal in a given year, and there are **101 cases**. Variables that I plan on using from this data set are `year`, `Total Deer`, and `Fall Turkey Harvest`, because these are the largest values and those that would be most relevant to the NICS data.

To import this table to RStudio, I first had to copy and paste the table into an Excel spreadsheet. I then cleaned the data by changing empty cells to "0" and removing all comma values. I also changed the variable names to have spaces removed. After cleaning the data a bit, I saved the Excel spreadsheet to my desktop as a CSV file, uploaded it to my Final Project Gihub repo, and used the `read_csv()` function to import the data table with the Gibhub link,  which ended up being successful. I saved this table as `PAHunting`.

```{r}
PAHunting_URL <- "https://raw.githubusercontent.com/ljbasom/Final-Project/master/PAHuntingData.csv"

PAHunting <- read_csv(PAHunting_URL)

head(PAHunting, 10)
```

My final data source is crime data from the FBI Uniform Crime Reporting Statistics (UCR) website. This division of the FBI collects crime data for each state on a year-by-year basis. The website has a helpful table builder tool that allows its users to select different aspects of the data to be put into a table that can be exported. For example, users can select any state or states that they desire, and different types of crime (or all types) that they want to be displayed in the data table. This data is collected and maintained by the FBI, and was collected to measure trends and changes in different types of crime across all states and the District of Columbia over a large span of years (**1960-2014**). I chose to create  tables that show crime data for all states and the District of Columbia (individually). A case in these tables that I created represents counts of different types of crimes for a certain year in the relevant state that the table is made for, and there are **55 cases** in each table (other than New York, which has **50 cases**), one for each year between 1960 and 2014 (and 1965 to 2014 for New York). Since all of these 51 tables were created separately (but shared the same variables), I was able to combine them within an Excel spreadsheet, add a new variable called `State`, and export this new combined table as a "csv" file to my desktop. When relating this data to the NICS data, I plan on using variables related to crime that would most likely be committed by a firearm. This would encompass variables like `Year`, `Population`, `Violent crime total` and then more specific variables related to certain types of violent crime.

To load the table into RStudio, I simply saved the combined table I created from the FBI UCR website to my desktop as a CSV file, uploaded it to my Final Project Github repo, and imported it the same way as I imported the PA hunting table, using `read_csv()`. I saved this table as `USCrime`.

```{r}
USCrime_URL <- "https://raw.githubusercontent.com/ljbasom/Final-Project/master/USCrime.csv"

USCrime <- read_csv(USCrime_URL)

head(USCrime, 10)
```


## Exploring intuition

In order to get a good grounding for this project and start on the right path, I need to do some preliminary plotting and observations to find direction and motivation for the work that I will be doing with these data sets in the near future. 

To start I want to look at the `NICS` data and filter it by `state`, using New Jersey to create graphs that begin to show trends in handgun permits by year for solely the NICS data:

```{r}
NICS %>%
  group_by(Year) %>%
  filter(state=="New Jersey") %>%
  summarise(Total = sum(handgun)) %>%
  ggplot(aes(x=Year, y=Total)) + geom_point() + theme(axis.text.x=element_text(angle=60,hjust=1))
```

Based on this nice graph, we can see that separating the `month` variable into `Month` and `Year` allowed us to make interesting plots when we group by the year. 

This plot shows us that the number of handgun permits was fairly steady from 1999 through about 2007, and then started to increase rapidly from 2008 through 2013, and then was kind of all over the place since then.

Another interesting thing we can realize from this plot is that the 2020 data point seems to be strangely low. That is because the year 2020 has data, but the year is not over yet, so the sum of handgun permits will naturally be lower in most cases. Until the 2020 data is complete for 12 months, we should probably refrain from including the year 2020 in creating plots like this.

For our next plot, we can look at the `USCrime` table to produce a plot of `Year` versus the proportion of violent crime to population (`Violent_crime_rate`) to make sure we account for increase in population, for 1999 through 2014 in New Jersey:


```{r}
USCrime %>%
  filter(State == "New Jersey") %>%
  ggplot(aes(x=as.integer(Year), y=Violent_Crime_rate)) + geom_point() + geom_smooth(span=.3)+ theme(axis.text.x=element_text(angle=60,hjust=1))
```

From this plot, we can see an interesting trend in the rate of violent crime in New Jersey from 1960 to 2014. While violent crime rates were on an upward trend for many years, they seemed to have peaked in 1990-1992 and have been consistently declining ever since. It will be interesting to see if the same trend will be observed within the NICS data for New Jersey, because the handgun data seems to follow a somewhat similar trend. It rises for a while, but instead of declning consistently, it seems to bounce between highs and lows for a while with numbers all over the place. This NICS handgun data does not take population into account, so we will be able to get a better understanding of trends once we can join the `NICS` and `USCrime` tables to have population become a bigger factor in the NICS data for New Jersey.

Let's now create a plot that shows the same information, only for the California crime data:

```{r}
USCrime %>%
  filter(State == "California") %>%
  ggplot(aes(x=as.integer(Year), y=Violent_Crime_rate)) + geom_point() + geom_smooth(span=.3)+ theme(axis.text.x=element_text(angle=60,hjust=1))
```

From the plot above, we can see a crime rate over the years that is very similar to the New Jersey crime rate plot. 

To make this comparison more clear and valid (instead of simply looking at two separate plots), we should make a plot that shows both New Jersey and California crime data, and create the same plot with color of the glyphs pertaining to the variable `State`:

```{r}
USCrime %>%
  filter(State == c("New Jersey", "California")) %>%
  ggplot(aes(x=as.integer(Year), y=Violent_Crime_rate)) + geom_point(aes(color=State)) + geom_smooth(span=.3, aes(color=State))+ theme(axis.text.x=element_text(angle=60,hjust=1))
```

This plot makes it easier to visualize and compare the violent crime rate over the years in both New Jersey and California. Since crime rate is simply the number of instances of crime divided by population, we can conclude that per capita, there is much more crime in California than there is in New Jersey, but both states had a crime rate peak (increase to decrease) around 1991-1993.

To go even further into analysis of crime rates, we can make the `USCrime` data narrow by creating two new columns, "Crime_Type" and "Crime_Rate" that contain the type of crime rate (violent crime, murder and nonnegligent manslaughter, legacy rape, robbery, aggravated assault, property crime, burglary, larceny theft and motor vehicle theft) and the rate value for each crime, in order to create a new plot that is still similar to the violent crime rate plot, but instead has all crime rates (for both states) contained in unique facets:

```{r}
CrimeRate <- 
  USCrime %>%
  pivot_longer(cols = c("Violent Crime" = Violent_Crime_rate, Murder_and_nonnegligent_manslaughter_rate, Legacy_rape_rate, Robbery_rate, Aggravated_assault_rate, Property_crime_rate, Burglary_rate, Larceny_theft_rate, Motor_vehicle_theft_rate), names_to = "Crime_Type", values_to = "Crime_Rate") %>%
  select(State, Year, Population, Crime_Type, Crime_Rate)
  
CrimeRate
```

```{r}
rate_labeller <- c(
  Violent_Crime_rate = "Violent Crime",
  Murder_and_nonnegligent_manslaughter_rate = "Murder",
  Legacy_rape_rate = "Legacy Rape",
  Robbery_rate = "Robbery",
  Aggravated_assault_rate = "Aggravated Assault",
  Property_crime_rate = "Property Crime",
  Burglary_rate = "Burglary",
  Larceny_theft_rate = "Larceny Theft",
  Motor_vehicle_theft_rate = "Vehicle Theft")

CrimeRate %>%
  filter(State == c("New Jersey", "California")) %>%
  ggplot(aes(x=Year,y=Crime_Rate))+aes(colour=State)+facet_wrap(~Crime_Type,ncol=4, scales = "free", labeller = labeller(Crime_Type = rate_labeller)) + stat_smooth(method=loess) + theme(axis.text.x=element_text(angle=60,hjust=1))
```

For our final plot in this intial observation of our data, we will look at the Pennsylvania hunting data. I want to see if the rates of deer and turkey harvest are rising and falling steadily. To do this, I will mutate a new variable that divides `Fall_Turkey_Harvest` by `Total_Deer` and plots this proportion over all years. If the line is flat, we can conclude that these numbers do tend to move together. If it slopes up, that means the number of turkeys goes up faster than the number of deer over the years, and if it slopes down, that means the number of deer goes up faster than the number of turkeys over the years:

```{r}
PAHunting %>% 
  mutate(proportion = Fall_Turkey_Harvest/Total_Deer) %>%
  ggplot(aes(x=Year, y=proportion)) + ylim(-1, 3) + geom_point()+ geom_smooth()+ theme(axis.text.x=element_text(angle=60,hjust=1))
```

This graph shows while at first the proportion was more strongly weighted by turkey (almost twice as many for a few years), this proportion evened out around 1930 and has stayed fairly consistent since then, meaning that the rates of success in turkey hunting and deer hunting have changed with one another since 1930.

## Closing remarks

I feel very confident after completing this stage of my final project. Although I have the challenge of working alone and doing everything by myself, I was able to successfully find three relevant data sets, one primary that I will be spending a majority of the analysis working on, and two supporting data sets to look into with the help of the primary data set.

Moving from this point, I have some more work to do. After completing many of the plots in the code above, I've realized that almost none of them pertain to the research question that I've asked, so I need to make sure I am refining the plots and visuals to answer certain aspects of my research question, or modify my research question to better fit the plots that I produced (especially in terms of US crime data). I also want to find ways to combine the hunting data and the gun permit data, as well as find ways to link the crime data with the gun permit data, and create visuals for these relationships.

Although my work in this stage of the project is somewhat cloudy and diverges from the main research question(s), I remain optimistic that I will easily be able to refine and clean-up my work by the time this project is done. I did some interesting things with R functions when making some of the plots like mutating variables, binding two tables together by their rows, narrowing the crime data by using crime type and rate, and creating faceted multi-variable plots that show important relationships in comparing crime rates between two states. I think with this level of data wrangling and visual creations, I should be able to stick with what I know (and what I have done in this code) and modify it to create plots that answer the initial research question more specifically.